First steps in synthetic algebraic geometry

### 2017-02-07 slides-warwick2017.tex
Final improvements

\begin{frame}\frametitle{Generic freeness}
  \vspace*{-1.6em}
  \[ \xymatrixcolsep{3pc}\xymatrixrowsep{3pc}\xymatrix{
    & \ar@{-}[d]^{\substack{\text{finitely}\\\text{generated}}} M \\
    R \ar[r]_{\text{of finite type}} & S
  } \]
  If $R$ is reduced ($x^n = 0 \Rightarrow x = 0$), there is a \hil{dense open subset} $U \subseteq \Spec R$ such that for any $f \in R$ with~$D(f)
  \subseteq U$,
  \begin{enumerate}
    \item $S[f^{-1}]$ and $M[f^{-1}]$ are free modules over $R[f^{-1}]$,
    \item $R[f^{-1}] \to S[f^{-1}]$ is of finite presentation, and
    \item $M[f^{-1}]$ is finitely presented as a module over~$S[f^{-1}]$.
  \end{enumerate}
  \pause

@@

\begin{frame}\frametitle{Synthetic algebraic geometry}
  Usual approach to algebraic geometry: \hil{layer schemes above ordinary set theory}
  using either
  \begin{itemize}
    \item locally ringed spaces
    \begin{multline*}
      \text{set of prime ideals of~$\ZZ[X,Y,Z]/(X^3+Y^3-Z^3)$} + {} \\
      \text{Zariski topology} + \text{structure sheaf}
    \end{multline*}
    \item or Grothendieck's functor-of-points approach, where a scheme is a functor~$\mathrm{Ring} \to \mathrm{Set}$.
    \[ A \longmapsto \{ (x,y,z) \in A^ 3 \,|\, x^3+y^3-z^3=0 \} \]
  \end{itemize}

  \hil{Synthetic approach}: model schemes \hil{directly as sets} in a certain
  nonclassical set theory.
  \[ \{ (x,y,z) \in (\affl)^3 \,|\, x^3+y^3-z^3=0 \} \]
\end{frame}

@@

\begin{frame}\frametitle{Synthetic constructions}
  \begin{itemize}
  \item $\mathbb{P}^n_S = \{ (x_0,\ldots,x_n) : (\affl)^{n+1} \,|\,
    x_0 \neq 0 \vee \cdots \vee x_n \neq 0 \}/(\affl)^\times.$ \\[1em]

  \item \ \\[-1.2em]\mbox{$\RelSpec R = \Hom_{\mathrm{Alg}(\affl)}(R, \affl) =
  \text{set of $\affl$-rational points of $R$}.$} \\[1em]

  \item $TX = \Hom(\Delta, X)$,
  where~$\Delta = \{ \varepsilon : \affl \,|\, \varepsilon^2 = 0 \}$.
  \end{itemize}
\end{frame}

@@

\begin{frame}{Synthetic formulation of properties}
  \begin{itemize}
  \item \ \\[-1.2em]\mbox{$\RelSpec R = \Hom_{\mathrm{Alg}(\affl)}(R, \affl) =
  \text{set of $\affl$-rational points of $R$}.$}

>>>>

\begin{frame}\frametitle{Generic freeness}
  \vspace*{-1.6em}
  \[ \xymatrixcolsep{3pc}\xymatrixrowsep{3pc}\xymatrix{
    & \ar@{-}[d]^{\substack{\text{finitely}\\\text{generated}}} M \\
    R \ar[r]_{\text{of finite type}} & S
  } \]
  If $R$ is reduced ($x^n = 0 \Rightarrow x = 0$), there
  %is a \hil{dense open subset} $U \subseteq \Spec R$ such that for any $f \in R$ with~$D(f)
  is $f \neq 0$ in $R$ such that
  \begin{enumerate}
    \item $S[f^{-1}]$ and $M[f^{-1}]$ are free modules over $R[f^{-1}]$,
    \item $R[f^{-1}] \to S[f^{-1}]$ is of finite presentation, and
    \item $M[f^{-1}]$ is finitely presented as a module over~$S[f^{-1}]$.
  \end{enumerate}
  \pause

@@

\begin{frame}\frametitle{Synthetic algebraic geometry}
  Usual approach to algebraic geometry: \hil{layer schemes above ordinary set theory}
  using either
  \begin{itemize}
    \item locally ringed spaces
    \small
    \begin{multline*}
      \text{set of prime ideals of~$\ZZ[X,Y,Z]/(X^n+Y^n-Z^n)$} + {} \\
      \text{Zariski topology} + \text{structure sheaf}
    \end{multline*}
    \normalsize
    \item or Grothendieck's functor-of-points account, where a scheme is a functor~$\mathrm{Ring} \to \mathrm{Set}$.
    \small\[ A \longmapsto \{ (x,y,z) \in A^3 \,|\, x^n+y^n-z^n=0 \} \]
  \end{itemize}
  \bigskip
  \pause

  \hil{Synthetic approach}: model schemes \hil{directly as sets} in a certain
  nonclassical set theory.
  \small
  \[ \{ (x,y,z) \in (\affl)^3 \,|\, x^n+y^n-z^n=0 \} \]
\end{frame}

@@

\begin{frame}\frametitle{Synthetic constructions}
  \begin{itemize}
  \item $\mathbb{P}^n = \{ (x_0,\ldots,x_n) : (\affl)^{n+1} \,|\,
    x_0 \neq 0 \vee \cdots \vee x_n \neq 0 \}/(\affl)^\times$ \\
  $\phantom{\mathbb{P}^n} \cong \text{set of one-dimensional subspaces of~$(\affl)^{n+1}$}$.

  \begin{itemize}
    \item $\O(1) = (\ell^\vee)_{\ell \in \mathbb{P}^n}$
    \item $\O(-1) = (\ell)_{\ell \in \mathbb{P}^n}$
    \item Euler sequence: $0 \to \ell^\perp \to ((\affl)^{n+1})^\vee \to \ell^\vee \to 0$
  \end{itemize}\bigskip

  \item \ \\[-1.2em]\mbox{$\RelSpec R = \Hom_{\mathrm{Alg}(\affl)}(R, \affl) =
  \text{set of $\affl$-valued points of $R$}.$}

  \begin{itemize}
    \item $\RelSpec \affl[X,Y,Z]/(X^n+Y^n-Z^n) \cong {}$ \\ \qquad\qquad $\{ (x,y,z) \in (\affl)^3 \,|\, x^n+y^n-z^n = 0 \}$ \medskip
    \item $\Delta \defeq \RelSpec \affl[\varepsilon]/(\varepsilon^2) \cong \{ \varepsilon : \affl \,|\, \varepsilon^2 = 0 \}$
  \end{itemize}\bigskip

  \item $TX = \Hom(\Delta, X)$.
  \end{itemize}
\end{frame}

@@

\begin{frame}{Synthetic formulation of properties}
  \begin{itemize}
  \item \ \\[-1.2em]\mbox{$\RelSpec R = \Hom_{\mathrm{Alg}(\affl)}(R, \affl) =
  \text{set of $\affl$-valued points of $R$}.$}
